Data Analysis Extensions

### 2015-10-15 RFCs/HDF5_Library/AnalysisExt/analysis_ext-body.tex
First bunch of revisions to analysis extensions
Former-commit-id: 6e2f23896444cf11519cdb913e399220ffd38344

\documentclass[letterpaper,hyper]{THG_RFC}

% Fonts
\usepackage{ifluatex}
\ifluatex
  % Font
  \usepackage{luatextra, xunicode, unicode-math}
  \defaultfontfeatures{Ligatures=TeX}
  \setmainfont[
    UprightFont={*},
    BoldFont={*-Bold},
    ItalicFont={*-Italic},
    BoldItalicFont={*-BoldItalic},
    SlantedFont={*-Italic},
    BoldSlantedFont={*-BoldItalic}
  ]{CMU Serif}
  \setsansfont[
    UprightFont={*},
    BoldFont={*-Bold},
    ItalicFont={*-Oblique},
    BoldItalicFont={*-BoldOblique}
  ]{CMU Sans Serif}
  \setmathfont[math-style=ISO, bold-style=ISO]{Asana Math}
  \setmonofont[
    UprightFont={*-Regular},
    BoldFont={*-Bold},
    ItalicFont={*-Italic},
    BoldItalicFont={*-BoldItalic}
  ]{CMU Typewriter}
\else
  \usepackage{times} % times / lmodern / mathpazo / palatino
  \usepackage[scaled=.95]{helvet}
  \usepackage{courier}
\fi

% Path to figures, plots
\graphicspath{{./pics/}{./plots/}}

%% TikZ
\usepackage{tikz}
\usetikzlibrary{patterns, shapes, decorations.pathreplacing}
\usepackage{gnuplot-lua-tikz}

% SI units
%\usepackage{textcomp}
\usepackage[binary-units]{siunitx}
\sisetup{per-mode = symbol}
\providecommand{\gbps}[1]{\SI{#1}{\giga\byte\per\second}}
\providecommand{\mbps}[1]{\SI{#1}{\mega\byte\per\second}}
\providecommand{\gb}[1]{\SI{#1}{\giga\byte}}
\providecommand{\mb}[1]{\SI{#1}{\mega\byte}}
\providecommand{\kb}[1]{\SI{#1}{\kilo\byte}}

% Code snippets
\usepackage{listings}
\def\lstsetc{\lstset{language=C,
  numbers=left,
  xleftmargin=20pt,
  numberstyle=\tiny\color{gray},
  stepnumber=1,
  showspaces=false, 
  showstringspaces=false,
  breaklines=true,
  basicstyle=\small\ttfamily,
  stringstyle=\itshape,
  commentstyle=\itshape\bfseries,
  morekeywords={main, size_t, malloc, free,
    hsize_t, hid_t, herr_t, H5X_class_t, H5X_type_t, H5Q_combine_op_t,
    H5Q_type_t, H5Q_match_op_t,
    H5Fcreate, H5Fclose,
    H5Dcreate, H5Dopen, H5Dclose, H5Dwrite, H5Dread, H5Dquery,
    H5Qcreate, H5Qclose, H5Qcombine, H5Qget_type, H5Qget_match_op,
    H5Qget_components, H5Qget_combine_op, H5Qencode, H5Qdecode,
    H5Screate_simple, H5Sclose, H5Sget_select_npoints,
    H5Xregister, H5Xunregister, H5Xcreate, H5Xremove, H5Xget_count,
    H5Vcreate, H5Vclose, H5Vget_location, H5Vget_query, H5Vget_counts,
    H5Vget_attrs, H5Vget_objs, H5Vget_elem_regions
    }
  }
}

\usepackage[margins]{trackchanges}
\addeditor{JS}
\addeditor{QK}
% annote, note, add, remove, change

% Title, author, etc
\title{Data Analysis Extensions}
\author{Jerome Soumagne}
\author{Quincey Koziol}
\date{July 24, 2015}
\rfcversion{2014-07-17.v4}
\revision{Jul. 17, 2014}{Version 1 circulated for comment within The HDF Group.}
\revision{Aug. 20, 2014}{Version 2 circulated for comment within The HDF Group.}
\revision{Nov. 14, 2014}{Version 3 circulated for comment within The HDF Group.}
\revision{Feb. 25, 2015}{Version 4 edits and typo fixing.}
\revision{Jul. 24, 2015}{Version 5 with edits to view object.}

%% Start the document
\begin{document}

%% Add Attribute reference RFC

%% Title
\maketitle

%% Abstract
\begin{abstract}
Accessing, selecting and retrieving data from an HDF5 container can be a time
consuming process, particularly so when data is very large. To enable, ease and 
accelerate this process, we introduce in this RFC extensions to the library
to efficiently query, select and index data.
\end{abstract}

@@

We define in this RFC the components that can enable application developers to
create complex and high-performance queries on both metadata and data elements
within an HDF5 container and retrieve the results of applying those query
operations. Support for these operations on HDF5 containers can be defined via:
\begin{enumerate}
\item New \textit{query} objects\footnote{Query and view objects are
\textit{in-memory} objects, which therefore do not modify the content of the
container.\label{fn:object}} and API routines, enabling the construction of
query requests for execution on HDF5 containers;
\item New \textit{view} objects\footref{fn:object} and API routines, which
\change[JS]{apply a query to an HDF5 container and return }{are the result of a query applied
to an HDF5 container and consist of} a set of references into the
container that fulfills the query criteria;
\item New \textit{index} objects and API routines, which allows the creation of
indices on the contents of HDF5 containers, to improve query performance.
\end{enumerate}

@@

The core query API is composed of
two routines: \texttt{H5Qcreate} and \texttt{H5Qcombine}. \texttt{H5Qcreate}
creates new queries, by specifying an aspect of an HDF5 container, such as
data elements, link names, attribute names, etc., a match operator, such as
$=$, $\neq$, $\leq$, $\geq$, and a value for the match operator.
Created query objects can be serialized and deserialized using \texttt{H5Qencode}
and \texttt{H5Qdecode} routines\footnote{Serialization/deserialization of queries
were introduced so that queries can be sent through the network.}, and their
content can be retrieved using the corresponding accessor routines.
\texttt{H5Qcombine} combines two query objects into a
new query object, using boolean operators such as $AND(\land)$ and $OR(\lor)$.
Queries created with \texttt{H5Qcombine} can be used as input to further
calls to \texttt{H5Qcombine}, creating more complex queries.

For example, a single call to \texttt{H5Qcreate} could create a query object
that would match data elements in any dataset within the container that are
equal to the value $17$. Another call to \texttt{H5Qcreate} could create a
query object that would match link names equal to $Pressure$.
Calling \texttt{H5Qcombine} with the $\land$ operator and those two query
objects would create a new query object that matched elements equal to $17$
in HDF5 datasets with link names equal to $Pressure$.
Creating the data analysis extensions to HDF5 using a \textit{programmatic
interface} for defining queries avoids defining a text-based query language
as a core component of the data analysis interface, and is more in keeping with
the design and level of abstraction of the HDF5 API.
The HDF5 data model is more complex than traditional database tables and a
simpler query model would likely not be able to express the kinds of queries
needed to extract the full set of components of an HDF5 container. A text-based
query language (or GUI) could certainly be built on top of the query API
defined here to provide a more user-friendly (as opposed to
\textit{developer-friendly}) query syntax like $Pressure = 17$. However, we
regard this as out-of-scope for now.

\TableRef{tab:querycomb} describes the result types for atomic queries and
combining queries of different types. Query results of $None$ type are rejected
when \texttt{H5Qcombine} is called, causing it to return failure\footnote{Query
results of $None$ type may be implemented with another result type in the
future, once experience with the query framework is acquired and a meaningful
grammar for those results are defined.}.

\begin{table}[ht]\footnotesize
\caption{Query combinations and associated result type.}
\label{tab:querycomb}
\begin{tabular}{ll} \toprule
Query &
Result Type\\ \midrule
\texttt{H5Q\_TYPE\_DATA\_ELEM} & $Dataset~Element$ \\
\texttt{H5Q\_TYPE\_ATTR\_VALUE} & $Attribute$ \\
\texttt{H5Q\_TYPE\_ATTR\_NAME} & $Object$ \\
\texttt{H5Q\_TYPE\_LINK\_NAME} & $Object$ \\
$Dataset~Element \land Dataset~Element$ & $Dataset~Element$ \\
$Dataset~Element \land Attribute$ & $None$ \\
$Dataset~Element \land Object$ & $Dataset~Element$ \\
$Attribute \land Attribute$ & $Attribute$ \\
$Attribute \land Object$ & $Attribute$ \\
$Object \land Object$ & $Object$ \\
$Dataset~Element \lor Dataset~Element$ & $Dataset~Element$ \\
$Dataset~Element \lor Attribute$ & $Combination$ \\
$Dataset~Element \lor Object$ & $Combination$ \\
$Dataset~Element \lor Combination$ & $Combination$ \\
$Attribute \lor Attribute$ & $Attribute$ \\
$Attribute \lor Object$ & $Combination$ \\
$Attribute \lor Combination$ & $Combination$ \\
$Object \lor Object$ & $Object$ \\
$Object \lor Combination$ & $Combination$ \\
$Combination \lor Combination$ & $Combination$ \\
$Combination \land Dataset~Element$ & $None$ \\
$Combination \land Attribute$ & $None$ \\
$Combination \land Object$ & $None$ \\
$Combination \land Combination$ & $None$ \\
\bottomrule
\end{tabular}
\end{table}

\section{View Objects}
Applying a query to an HDF5 container creates an HDF5 view. HDF5 view
objects are run-time, in-memory objects (i.e., not stored in a container
\add[JS]{but which can be persisted}), that
consist of read-only references \add[JS]{to attributes, objects and data regions}
into the contents of the HDF5 container that the query was applied to.
\add[JS]{The view object can be therefore seen as a virtual container. It is
defined as an anonymous HDF5 group, which consists of three predefined datasets that
contain attribute, object and region references, and stored in-memory, i.e., using
the existing core driver. A new view is created by the
following routine, which applies a query to an HDF5 container, group hierarchy,
or individual object and returns the object ID of the newly created group:}
{\lstsetc
\begin{lstlisting}
hid_t H5Qapply(hid_t loc_id, hid_t query_id, unsigned *result, hid_t vcpl_id);
\end{lstlisting}
}

\add[JS]{As mentioned, the created view may also need to be persisted and this
can be done by calling \textit{H5Ocopy()} to copy the group (stored in a virtual
container) to a persistent container. For
coherency, a time stamp may be attached to it so that its states has a meaning
compared to the state of the container that the query was applied to (as the
container may have been modified in the meantime). It may also be useful in
that case to define different states to the view (\textit{dead} or
\textit{live}) so that the user knows whether the view is current or not.}

@@

\remove[JS]{View objects are created with \texttt{H5Vcreate}, which applies a query to an
HDF5 container, group hierarchy, or individual object and produces the view
object as a result. The attributes, objects, and/or data elements referenced
within a view can be retrieved by further API calls.} Note that currently,
attribute references are not available, this feature will be added in order to
support views that contain attribute references\footnote{See RFC 201X-XX-XX.vX.}.

\begin{figure}
\input{pics/hdf5_view1}
\caption{HDF5 container example.}
\label{fig:hdf5_view1}
\end{figure}

For example, starting with the HDF5 container described in~\FigureRef{fig:hdf5_view1},
applying the $link\_name = Pressure$ query (described above) would result
in the view shown in~\FigureRef{fig:hdf5_view2}, highlighted in blue.

\begin{figure}
\input{pics/hdf5_view2}
\caption{HDF5 container example with query $link\_name = Pressure$ applied.}
\label{fig:hdf5_view2}
\end{figure}

Alternatively, applying the $data\_element = 17$ query (described above) would
result in the view shown in~\FigureRef{fig:hdf5_view3}, highlighted in blue.

\begin{figure}
\input{pics/hdf5_view3}
\caption{HDF5 container example with query $data\_element = 17$ applied.}
\label{fig:hdf5_view3}
\end{figure}

Finally, applying the combined $(link\_name = Pressure)\land(data\_element = 17)$
query (described above) would result in the view shown in~\FigureRef{fig:hdf5_view4},
highlighted in blue.

\begin{figure}
\input{pics/hdf5_view4}
\caption{HDF5 container example with query $(link\_name = Pressure)\land(data\_element = 17)$ applied.}
\label{fig:hdf5_view4}
\end{figure}

Views can be thought of as containing a set of HDF5 references (object,
dataset region or attribute references) to components of the underlying
container, retaining the context of the original container. For example, the
view containing the results of the $(link\_name = Pressure)\land(data\_element = 17)$
query will contain three dataset region references, which
can be retrieved from the view object and probed for the dataset and selection
containing the elements that match the query with the existing \texttt{H5Rdereference}
and \texttt{H5Rget\_region} API calls. Note that selections returned from a region
reference retain the underlying dataset's dimensionality and coordinates---they
are not \textit{flattened} into a 1-D series of elements. The selection returned
from a region reference can also be applied to a different dataset in the container,
allowing a query on pressure values to be used to extract temperature values,
for example.

\section{Index Objects}
Index objects are designed to accelerate creation of view objects from
frequently occurring query operations.
For example, if the $(link\_name = Pressure)\land(data\_element = 17)$ query
(previously described) is going to be frequently executed on the container, indices
could be created in that container, which would speed up the creation of views
when querying for link names and for data element values. Indices created for
accelerating the $link\_name = Pressure$ or $data\_element = 17$ queries
would also improve view creation for the more complex
$(link\_name = Pressure)\land(data\_element = 17)$ query.

Although creating indices for metadata components of queries, such as link or
attribute names, is possible, we focus on index creation for dataset elements,
as they represent the largest volume of data in typical HPC application usage of
HDF5. Queries with metadata components execute properly,
\todo{Metadata indexing must be added.} but are not able to be accelerated
with an index currently.

The indexing API can work in conjunction with the view API. When an \texttt{H5Vcreate}
call is made for a group or dataset, an index attached to any dataset queried
for element value ranges will be used to speed up the query process and return
a dataspace selection to the library for later use.

There are different techniques for creating data element indices, and the most
efficient method will vary depending on the type of the data that is to be
indexed, its layout, etc. A new interface for the HDF5 library that uses a
plugin mechanism is therefore defined.

@@

Index objects are stored in the HDF5 container that they apply to, but are not
visible in the container's group hierarchy\footnote{Plugin developers, note that
the HDF5 library's existing anonymous dataset and group creation calls can be
used to create objects in HDF5 files that are not visible in the container's
group hierarchy.}.
Instead, index objects are part of the metadata for the file itself. New index
objects are created by passing an HDF5 container to be indexed and the index
plugin ID to the \texttt{H5Xcreate} call.
Alternatively an index may be created at the same time as a dataset gets created
by passing a property to the dataset creation property list.
Index information (such as plugin ID and index metadata) is stored at index
creation time\footnote{Adding index information introduces a file format change.},
and when the user later calls \texttt{H5Dopen}, the plugin open
callback will retrieve this stored information and make use of the corresponding
index plugin for all subsequent operations. Similarly, calling \texttt{H5Dclose}
will call the plugin index close callback and close the objects used to store
the index data.

\begin{figure}
\input{pics/hdf5_index}
\caption{Index information (plugin ID and metadata) is stored along the object header.}
\label{fig:hdf5_index}
\end{figure}

@@

When a call to \texttt{H5Vcreate} is made, the index plugin query callback will be
invoked to create a selection of elements in the dataset that match the query
parameters. Applications can also use the new \texttt{H5Dquery} routine defined below
to directly
execute a query on a particular dataset (accelerated by any index defined on the dataset),
and retrieve the selection that matches the query.

@@

Because the amount of space taken by the index cannot be directly retrieved by
the user (since the datasets storing the indices are known only by the plugin
itself), the \texttt{get\_size} callback can query the amount of space that
the index takes in the file and users may query that information using the
corresponding \texttt{H5Xget\_size} routine.

@@

Implementations for FastBit and ALACRITY index packages are already
present, as well as a brute force indexing plugin. Early performance results
are presented in~\FigureRef{fig:indexing_perf}.

\begin{figure}
\input{plots/indexing_perf}
\caption{Indexing performance.}
\label{fig:indexing_perf}
\end{figure}

@@

\subsubsection*{Questions}
Some questions are still open regarding the handling of indices:
\begin{enumerate}
%\item How could an index be built and queried in parallel for a dataset that already exists?
\item How to handle index updates when the specified index plugin is not available?
(In traditional databases, stored procedures are saved with the data and
therefore available at any time, but that is not the case here)
We could mark the index as out of date and rebuild the index when the plugin is
available again.
\item How to handle index queries when the specified index plugin is not
available? We could fallback to another plugin and do a brute force query on the data.
\end{enumerate}

@@

When the dataset is opened and the index reopened, we can lookup the index
information in the B-tree that corresponds to each chunk and call the \texttt{open}
callback using the associated metadata.

%write
%   if (chunked)
%     modify I/O loop

Similarly, when a query is issued and needs to be answered, the chunks that
correspond to the selection passed to the \texttt{H5Dquery} call are selected
and their index is used to answer that query. The selection returned is then
added to a global selection, which is then in turn returned to the user.

Finally, when \texttt{H5Xremove} is called, the \texttt{delete} plugin callback is
invoked for each chunk, by using the index metadata information stored in the B-tree.

@@

An important design choice to support parallel indexing is to give as much
freedom as possible to the indexing plugin developer so that in the case when
the indexing library supports parallel indexing, it is still possible to take
advantage of it. Three options are available to support parallel indexing
from the previous interface:
\begin{enumerate}
\item Let the index creation be collective. This however implies having synchronization
points, which is the main constraint.
\item Let the index creation be independent. However, creating datasets to
store the index data must be done collectively\footnote{An option could be to use
the metadata server VOL plugin but this option is not easily doable yet.}.
\item Make the index creation in two phases. One that consists of
building the index in parallel, independently, and gathering the information (index size, etc)
at the end. The other that consists of letting the dataset creation be done by a master
process, which can then let the other processes write the index data independently.
There could however be a memory constraint in this case if the index data has to
be kept in memory between these two phases (though building the index twice is not
a good solution either).
\end{enumerate}

@@

\section{Conclusion}
Document in progress.

\section*{Revision History}
\makerevisions

%% References
\bibliographystyle{ieeetr}
\bibliography{analysis_ext}

\end{document}

>>>>

@@

We define in this RFC the components that can enable application developers to
create complex and high-performance queries on both metadata and data elements
within an HDF5 container and retrieve the results of applying those query
operations. Support for these operations can be defined via:
\begin{itemize}
\item New \textit{query} objects\footnote{Query objects are
\textit{in-memory} objects, which therefore do not modify the content of the
container.\label{fn:object}} and API routines, enabling the construction of
query requests for execution on HDF5 containers;
\item New \textit{index} objects and API routines, which allows the creation of
indices on the contents of HDF5 containers, to improve query performance.
\end{itemize}

@@

\subsection{Query Creation}

The core query API is composed of
two routines: \texttt{H5Qcreate} and \texttt{H5Qcombine}. \texttt{H5Qcreate}
creates new queries, by specifying an aspect of an HDF5 container, such as:
\begin{itemize}
\item \texttt{H5Q\_TYPE\_DATA\_ELEM} (data element);
\item \texttt{H5Q\_TYPE\_LINK\_NAME} (link name);
\item \texttt{H5Q\_TYPE\_ATTR\_NAME} (attribute name);
\item \texttt{H5Q\_TYPE\_ATTR\_VALUE} (attribute value);
\end{itemize}
as well as a match operator, such as:
\begin{itemize}
\item \texttt{H5Q\_MATCH\_EQUAL} ($=$);
\item \texttt{H5Q\_MATCH\_NOT\_EQUAL} ($\neq$);
\item \texttt{H5Q\_MATCH\_LESS\_THAN} ($\leq$);
\item \texttt{H5Q\_MATCH\_GREATER\_THAN} ($\geq$);
\end{itemize}
and a value for the match operator.
Created query objects can be serialized and deserialized using the \texttt{H5Qencode}
and \texttt{H5Qdecode} routines\footnote{Serialization/deserialization of queries
were introduced so that queries can be sent through the network.}, and their
content can be retrieved using the corresponding accessor routines.
\texttt{H5Qcombine} combines two query objects into a
new query object, using boolean operators such as:
\begin{itemize}
\item \texttt{H5Q\_COMBINE\_AND} ($\mathrm{\land}$);
\item \texttt{H5Q\_COMBINE\_OR} ($\mathrm{\lor}$).
\end{itemize}
Queries created with \texttt{H5Qcombine} can be used as input to further
calls to \texttt{H5Qcombine}, creating more complex queries.

For example, a single call to \texttt{H5Qcreate} could create a query object
that would match data elements in any dataset within the container that are
equal to the value ``$\mathrm{17}$''. Another call to \texttt{H5Qcreate} could create a
query object that would match link names equal to ``$\mathrm{Pressure}$''.
Calling \texttt{H5Qcombine} with the $\mathrm{\land}$ operator and those two query
objects would create a new query object that matched elements equal to ``$\mathrm{17}$''
in HDF5 datasets with link names equal to ``$\mathrm{Pressure}$''.
Creating the data analysis extensions to HDF5 using a \textit{programmatic
interface} for defining queries avoids defining a text-based query language
as a core component of the data analysis interface, and is more in keeping with
the design and level of abstraction of the HDF5 API.
The HDF5 data model is more complex than traditional database tables and a
simpler query model would likely not be able to express the kinds of queries
needed to extract the full set of components of an HDF5 container. A text-based
query language (or GUI) could certainly be built on top of the query API
defined here to provide a more user-friendly (as opposed to
\textit{developer-friendly}) query syntax like ``$\mathrm{Pressure = 17}$''.
However, we regard this as out-of-scope for now.

\subsection{Query Execution and Views}
Applying a query to an HDF5 container creates an HDF5 \textit{view}. A view is
an anonymous HDF5 group in a temporary memory container. A view group contains up to
three 1-D dataset objects with references into the contents of the HDF5 container
that the query was applied to:
one with \textit{object} references, one with \textit{region} references and
one with \textit{attribute} references.

\begin{figure}
\input{pics/hdf5_view}
\caption{Temporary memory container with datasets containing query results (i.e.,
HDF5 references).}
\label{fig:hdf5_view}
\vspace{-20pt}
\end{figure}

A new view is created by the
following \texttt{H5Qapply} routine, which applies a query to an HDF5 container,
group hierarchy, or individual object and returns the object ID of the newly created
group. The attributes, objects, and/or data regions referenced within a view's
datasets can be retrieved by further HDF5 dataset (H5D*) API calls.

{\lstsetc
\begin{lstlisting}
hid_t H5Qapply(hid_t loc_id, hid_t query_id, unsigned *result, hid_t vcpl_id);
\end{lstlisting}
}

Although the created view is stored in memory, it can be persisted
by calling \texttt{H5Ocopy()} to copy the group (stored in a virtual
container) to a persistent container. For
coherency, a time stamp may also be attached to it so that its states has a meaning
compared to the state of the container that the query was applied to (as the
container may have been modified in the meantime). It may be useful in
that case to define different states to the view (\textit{dead} or
\textit{live}) so that the user knows whether the view is current or not.

@@

%Note that currently, attribute references are not available, this feature will be added in order to
%support views that contain attribute references\footnote{See RFC 201X-XX-XX.vX.}.

For example, starting with the HDF5 container described in~\FigureRef{fig:hdf5_view1},
applying a ``$\mathrm{link\_name}$ $\mathrm{=}$ $\mathrm{Pressure}$''
query would result in the view shown in~\FigureRef{fig:hdf5_view2}, highlighted in blue.
\begin{figure}
\begin{subfigure}[b]{.49\linewidth}
\centering
\resizebox{\textwidth}{!}{
  \input{pics/hdf5_view1}
}
\caption{HDF5 container example.}
\label{fig:hdf5_view1}
\end{subfigure}
\hfill
\begin{subfigure}[b]{.49\linewidth}
\centering
\resizebox{\textwidth}{!}{
  \input{pics/hdf5_view2}
}
\caption{HDF5 container with link name query applied.}
\label{fig:hdf5_view2}
\end{subfigure}
\caption{HDF5 view creation example with simple query.}
\end{figure}

Alternatively, applying a ``$\mathrm{data\_element}$ $\mathrm{=}$ $\mathrm{17}$''
query would result in the view shown in~\FigureRef{fig:hdf5_view3}, highlighted in blue.
Finally, applying the combined queries ``$\mathrm{(link\_name = Pressure)}$
$\mathrm{\land}$ $\mathrm{(data\_element = 17)}$''
would result in the view shown in~\FigureRef{fig:hdf5_view4}, highlighted in blue.

\begin{figure}
\begin{subfigure}[b]{.49\linewidth}
\centering
\resizebox{\textwidth}{!}{
  \input{pics/hdf5_view3}
}
\caption{HDF5 container with data element query applied.}
\label{fig:hdf5_view3}
\end{subfigure}
\hfill
\begin{subfigure}[b]{.49\linewidth}
\centering
\resizebox{\textwidth}{!}{
  \input{pics/hdf5_view4}
}
\caption{HDF5 container with combined query applied.}
\label{fig:hdf5_view4}
\end{subfigure}
\caption{HDF5 view creation example with combined query.}
\vspace{-20pt}
\end{figure}

Since views contain a set of HDF5 references (object,
dataset region or attribute references) to components of the underlying
container, they retain the context of the original container. For example, the
view containing the results in~\FigureRef{fig:hdf5_view4} will contain three
dataset region references, which
can be retrieved from the view object and probed for the dataset and selection
containing the elements that match the query with the existing \texttt{H5Rdereference}
and \texttt{H5Rget\_region} API calls. Note that selections returned from a region
reference retain the underlying dataset's dimensionality and coordinates---they
are not \textit{flattened} into a 1-D series of elements. The selection returned
from a region reference can also be applied to a different dataset in the container,
allowing a query on pressure values to be used to extract temperature values,
for example.

\TableRef{tab:querycomb} describes the result types for atomic queries and
combining queries of different types. Query results of \textit{None} type are rejected
when \texttt{H5Qcombine} is called, causing it to return failure\footnote{Query
results of \textit{None} type may be implemented with another result type in the
future, once experience with the query framework is acquired and a meaningful
grammar for those results are defined.}.

\begin{table}[ht]\footnotesize
\caption{Query combinations and associated result reference type.}
\label{tab:querycomb}
\begin{tabular}{ll} \toprule
Query &
Result Type\\ \midrule
\texttt{H5Q\_TYPE\_DATA\_ELEM} & Dataset~Region \\
\texttt{H5Q\_TYPE\_ATTR\_VALUE} & Attribute \\
\texttt{H5Q\_TYPE\_ATTR\_NAME} & Attribute \\
\texttt{H5Q\_TYPE\_LINK\_NAME} & Object \\
Dataset~Element $\mathrm{\land}$ Dataset~Element & Dataset~Region \\
Dataset~Element $\mathrm{\land}$ Attribute & Dataset~Region \\
Dataset~Element $\mathrm{\land}$ Object & Dataset~Region \\
Attribute $\mathrm{\land}$ Attribute & Attribute \\
Attribute $\mathrm{\land}$ Object & Object \\
Object $\mathrm{\land}$ Object & Object \\
Dataset~Element $\mathrm{\lor}$ Dataset~Element & Dataset~Region \\
Dataset~Element $\mathrm{\lor}$ Attribute & Combination \\
Dataset~Element $\mathrm{\lor}$ Object & Combination \\
Dataset~Element $\mathrm{\lor}$ Combination & Combination \\
Attribute $\mathrm{\lor}$ Attribute & Attribute \\
Attribute $\mathrm{\lor}$ Object & Combination \\
Attribute $\mathrm{\lor}$ Combination & Combination \\
Object $\mathrm{\lor}$ Object & Object \\
Object $\mathrm{\lor}$ Combination & Combination \\
Combination $\mathrm{\lor}$ Combination & Combination \\
Combination $\mathrm{\land}$ Dataset~Element & None \\
Combination $\mathrm{\land}$ Attribute & None \\
Combination $\mathrm{\land}$ Object & None \\
Combination $\mathrm{\land}$ Combination & None \\
\bottomrule
\end{tabular}
\end{table}

\section{Index Objects}
Index objects are designed to accelerate creation of views from query operations.
For example, if the ``$\mathrm{(link\_name = Pressure)}$
$\mathrm{\land}$ $\mathrm{(data\_element = 17)}$'' query
(previously described) is going to be either frequently executed or executed
on a container that contains significantly large datasets, indices
could be created in that container, which would speed up the
creation of views when querying for link names and for data element values.
Indices created for accelerating the $link\_name = Pressure$ or $data\_element = 17$ queries
would also improve view creation for the more complex
$(link\_name = Pressure)\land(data\_element = 17)$ query.
Although creating indices for metadata components of queries, such as link or
attribute names, is possible, we first focus on index creation for dataset elements,
as they represent the largest volume of data in typical HPC application usage of
HDF5. Queries with metadata components execute properly,
but are not able to be accelerated with an index currently.

The indexing API can work in conjunction with the view creation. When an \texttt{H5Qapply}
call is made on a given location, an index attached to any dataset queried
for element value ranges will be used to speed up the query process and return
a dataspace selection to the library for later use.
There are different techniques for creating data element indices, and the most
efficient method will vary depending on the type of the data that is to be
indexed, its layout, etc. A new interface for the HDF5 library that uses a
plugin mechanism is therefore defined.

@@

Index objects are stored in the HDF5 container that they apply to, but are not
visible in the container's group hierarchy\footnote{Plugin developers, note that
the HDF5 library's existing anonymous dataset and group creation calls can be
used to create objects in HDF5 files that are not visible in the container's
group hierarchy.}.
Instead, index objects are part of the metadata for the indexed dataset. New index
objects are created by passing an HDF5 location (group or dataset) to be indexed
and the index plugin ID to the \texttt{H5Xcreate} call.
Alternatively an index may be created at the same time as a dataset gets created
by passing a property to the dataset creation property list.
Index information (such as plugin ID and index metadata) is stored at index
creation time\footnote{Adding index information introduces a file format change.},
and when the user later calls \texttt{H5Dopen}, the plugin open
callback will retrieve this stored information and make use of the corresponding
index plugin for all subsequent operations\footnote{Note that the dataset's
index object is opened only when a query call is made on the dataset.}.
Similarly, calling \texttt{H5Dclose}
will call the plugin index close callback and close the objects used to store
the index data.

\begin{figure}
\input{pics/hdf5_index}
\caption{Index information (plugin ID and metadata) is stored along the object header.}
\label{fig:hdf5_index}
\vspace{-20pt}
\end{figure}

@@

When a call to \texttt{H5Qapply} is made, the index plugin query callback will be
invoked to create a selection of elements in the dataset that match the query
parameters. Applications can also directly use the internally called
\texttt{H5Dquery} routine defined below
to directly execute a query on a particular dataset (accelerated by any index defined on the dataset),
and retrieve the selection that matches the query.

@@

Because the amount of space taken by the index cannot be directly retrieved by
the user (since the datasets storing the indices are known only by the plugin
itself), the \texttt{get\_size} callback can query the amount of space that
the index takes in the file and users may correspondingly query that information using the
\texttt{H5Xget\_size} routine.

@@

Implementations for FastBit and ALACRITY index packages are already
present, as well as a dummy brute force indexing plugin. Early performance results
are presented in~\FigureRef{fig:indexing_perf}.

\begin{figure}
\input{plots/indexing_perf}
\caption{Indexing performance.}
\label{fig:indexing_perf}
\vspace{-20pt}
\end{figure}

@@

\subsubsection*{Questions}
Some questions are still open regarding the handling of indices:
\begin{enumerate}
%\item How could an index be built and queried in parallel for a dataset that already exists?
\item How to handle index updates when the specified index plugin is not available?
\textit{In traditional databases, stored procedures are saved with the data and
therefore available at any time, but that is not the case here.
We could mark the index as out of date and rebuild the index when the plugin is
available again.}
\item How to handle index queries when the specified index plugin is not
available? \textit{We could fallback to another plugin and do a brute force query on the data.}
\end{enumerate}

@@

When the dataset is opened and the index reopened, we can lookup the index
information in the B-tree that corresponds to each chunk and call the \texttt{open}
callback using the associated metadata.
%write
%   if (chunked)
%     modify I/O loop
Similarly, when a query is issued and needs to be answered, the chunks that
correspond to the selection passed to the \texttt{H5Dquery} call are selected
and their index is used to answer that query. The selection returned is then
added to a global selection, which is then in turn returned to the user.
Finally, when \texttt{H5Xremove} is called, the \texttt{delete} plugin callback is
invoked for each chunk, by using the index metadata information stored in the B-tree.

@@

An important design choice to support parallel indexing is to give as much
freedom as possible to the indexing plugin developer so that in the case when
the indexing library supports parallel indexing, it is still possible to take
advantage of it. Three options are available to support parallel indexing
from the previous interface:
\begin{itemize}
\item Let the index creation be collective. This however implies having synchronization
points, which is the main constraint.
\item Let the index creation be independent. However, creating datasets to
store the index data must be done collectively\footnote{An option could be to use
the metadata server VOL plugin but this option is not easily doable yet.}.
\item Make the index creation in two phases. One that consists of
building the index in parallel, independently, and gathering the information (index size, etc)
at the end. The other that consists of letting the dataset creation be done by a master
process, which can then let the other processes write the index data independently.
There could however be a memory constraint in this case if the index data has to
be kept in memory between these two phases (though building the index twice is not
a good solution either).
\end{itemize}

@@

%\section{Conclusion}
%Document in progress.
